\documentclass[11pt]{article}

\usepackage{fullpage}
\usepackage{times}
\usepackage{hyperref,microtype,pdfsync}
\usepackage{amsmath,amsfonts,amssymb,amsthm}
\usepackage{mathtools}
\usepackage{fancyhdr}

\input{header}

% VARIABLES

\newcommand{\lecturenum}{19}
\newcommand{\lecturetopic}{ZK Potpourri}
\newcommand{\scribename}{YOUR NAME HERE}

% END OF VARIABLES

\lecheader

\pagestyle{plain}              % default: no special header

\begin{document}

\thispagestyle{fancy}          % first page should have special header

% LECTURE MATERIAL STARTS HERE

\section{Recap}
\label{sec:recap}

Relation $R$.  ZK with efficient prover for a relation.

\section{Witness Indistinguishability}
\label{sec:witn-indist}

A relaxation of ZK and its uses.

\subsection{Definition}
\label{sec:definition}

\begin{definition}
  \label{def:wi}
  An efficient prover $P(\cdot,\cdot)$ for a relation $R$ is
  \emph{witness indistinguishable} if for all nuppt $V^{*}$, for all
  $x \in L_{R}$ and any $w_{0}, w_{1} \in R(x)$, \[
  \view_{V^{*}}[P(x,w_{0}) \leftrightarrow V^{*}(x)] \compind
  \view_{V^{*}}[P(x,w_{1}) \leftrightarrow V^{*}(x)]. \]
\end{definition}

\begin{lemma}
  \label{lem:zk-implies-wi}
  Let $(P,V)$ be a proof system (with efficient prover) for a relation
  $R$.  If $(P,V)$ is zero-knowledge, then it is
  witness-indistinguishable.
\end{lemma}

\begin{proof}
  TODO.
\end{proof}

Is the converse of Lemma~\ref{lem:zk-implies-wi} true?  That is, does
witness indistinguishability imply zero-knowledge?

\subsection{Properties}
\label{sec:properties}

\begin{lemma}
  \label{lem:wi-parallel}
  Witness indistinguishability is preserved under parallel repetition.
\end{lemma}

\begin{corollary}
  \label{cor:3col-parallel-wi}
  The protocol for $\problem{3COL}$ repeated $n^{3}$ times in parallel
  is a $3$-round witness-indistinguishable protocol having soundness
  error at most $2^{-n}$.
\end{corollary}

\begin{proof}[Proof of Lemma~\ref{lem:wi-parallel}]
  TODO.
\end{proof}

\subsection{Applications}
\label{sec:applications}

TODO: describe high-level idea behind $O(1)$-round protocol (with
$\negl(n)$ soundness error) for any $\NP$ language.

\section{Noninteractive Zero Knowledge}
\label{sec:nonint-zero-knowl}



\end{document}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:
