\documentclass[11pt]{article}

\usepackage{fullpage}
\usepackage{mathptmx}
\usepackage[T1]{fontenc}
\usepackage{hyperref,microtype,pdfsync}
\usepackage{amsmath,amsfonts,amssymb,amsthm}
\usepackage{fancyhdr}
\usepackage{mathtools}
\usepackage{algorithm,algorithmic}

\input{header}

% VARIABLES

\newcommand{\lecturenum}{3}
\newcommand{\lecturetopic}{OWFs, Hardness Amplification}
\newcommand{\scribename}{George P.~Burdell}

% END OF VARIABLES

\lecheader

\pagestyle{plain}               % default: no special header

\begin{document}

\thispagestyle{fancy}           % first page should have special header

% LECTURE MATERIAL STARTS HERE

\section{One-Way Functions}
\label{sec:one-way-functions}

Recall our formal definition of a one-way function:
\begin{definition}
  \label{def:owf}
  A function $f : \bit^{*} \to \bit^{*}$ is \emph{one-way} if it
  satisfies the following conditions.
  \begin{itemize}
  \item \emph{Easy to compute.}  There is an efficient algorithm
    computing $f$.  Formally, there exists a (uniform, deterministic)
    poly-time algorithm $\algo{F}$ such that $\algo{F}(x) = f(x)$ for
    all $x \in \bit^{*}$.
  \item \emph{Hard to invert.}  An efficient algorithm inverts $f$ on
    a random input with only negligible probability.  Formally, for
    any non-uniform PPT algorithm $\Inv$, the \emph{advantage} of
    $\Inv$ \[ \advan_{f}(\Inv) := \Pr_{x \gets \bit^{n}} \bracks*{
      \Inv(1^{n}, f(x)) \in f^{-1}(f(x)) } = \negl(n) \] is a
    negligible function (in the security parameter $n$).
  \end{itemize}
\end{definition}
We also allowed a syntactic relaxation where the domain and range (and
the input distribution over the domain) may be arbitrary, as long as
the input distribution can be sampled efficiently (given the security
parameter $n$).

\subsection{Candidates}
\label{sec:candidates}

At the end of last lecture we defined two functions, with the idea of
investigating whether they could be one-way functions:
\begin{enumerate}
\item Subset-sum: define $f_{\text{ss}} : (\Z_{N})^{n} \times \bit^{n}
  \to (\Z_{N})^{n} \times \Z_{N}$, where $N = 2^{n}$, as \[
  f_{\text{ss}}(a_{1}, \ldots, a_{n}, b_{1}, \ldots, b_{n}) = (a_{1},
  \ldots, a_{n}, S = \sum_{i\; :\; b_{i} = 1} a_{i} \bmod N). \]
  Notice that because the $a_{i}$s are part of the output, inverting
  $f$ means finding a subset of $\set{1,\ldots,n}$ (corresponding to
  those $b_{i}$s equaling $1$) that induces the given sum $S$ modulo
  $N$.
\item Multiplication: define $f_{\text{mult}} : \N^{2} \to \N$
  as\footnote{For security parameter $n$, we use the domain $[1,2^{n}]
    \times [1,2^{n}]$.  The cases $x=1$ and $y=1$ are special to rule
    out the trivial preimages $(1,xy)$ and $(xy, 1)$ for the output
    $xy \neq 1$.}
  \[
  f_{\text{mult}}(x,y) =
  \begin{cases}
    1 & \text{if } x=1 \vee y=1 \\
    x \cdot y & \text{otherwise.}
  \end{cases}
  \]
\end{enumerate}

First consider $f_{\text{ss}}$; clearly it can be computed in
polynomial time.  The subset-sum problem is a famous $\NP$-complete
problem, hence if $\P \neq \NP$ there is no poly-time algorithm to
solve it, and $f_{\text{ss}}$ must be one-way, right?  Not exactly:
subset-sum is $\NP$-complete in the \emph{worst case}, but for
$f_{\text{ss}}$ to be one-way we need it be hard on the
\emph{average}.  Nevertheless, many believe that this form of
subset-sum is indeed average-case hard, and hence $f_{\text{ss}}$ is
conjectured to be one-way.

Now consider $f_{\text{mult}}$; again, it can be computed in
poly-time.  Inverting $f_{\text{mult}}$ means factoring $xy$ (into
non-trivial factors), and factoring is hard, right?  Again, we need to
be more careful: consider the following inverter $\Inv(z)$: if $z$ is
even, then output $(2, z/2)$, otherwise fail.  Notice that $\Inv$
outputs a valid preimage of $z$ exactly when $z = xy$ is even, which
occurs with probability $3/4$ when $x, y$ are chosen independently and
uniformly from $[1,2^{n}]$.

The problem here is that while \emph{some} natural numbers seem hard
to factor, a large fraction of them (e.g., those that are even)
certainly are not.  There are two common ways to deal with this.  The
first is to ``tweak'' the function (actually, its input distribution)
to focus on the instances that we believe to be hard.  For the
factoring problem, it is believed (but not proved!) that the hardest
numbers to factor are those that are the product of exactly two primes
of about the same size.  More specifically, define \[ \Pi_{n} = \set{
  p\; :\; p \in [1,2^{n}] \text{ is prime}} \] to be the set of
``$n$-bit prime numbers.''  Then we may make the following plausible
conjecture:

\begin{conjecture}[Factoring Assumption]
  \label{conj:fact-primes-hard}
  For every non-uniform PPT algorithm $\Adv$,
  \[ \Pr_{p,q \gets \Pi_{n}} [\Adv(p \cdot q) = (p,q)] = \negl(n). \]
\end{conjecture}

Let's next change the input distribution of $f_{\text{mult}}$ so that
$x,y$ are chosen uniformly and independently from $\Pi_{n}$.  Now, the
one-wayness property for $f_{\text{mult}}$ is syntactically identical
to Conjecture~\ref{conj:fact-primes-hard}, so for that property we
have a ``proof by assumption.''  The only remaining question is the
distribution over inputs: can we efficiently sample a random $n$-bit
prime, i.e., sample uniformly from $\Pi_{n}$?
Algorithm~\ref{alg:genprime} gives a potential algorithm,
$\algo{GenPrime}$, to do so.

\begin{algorithm}
  \caption{Algorithm $\algo{GenPrime}(1^{n})$ for sampling a uniformly
    random $n$-bit prime.}
  \label{alg:genprime}
  \begin{algorithmic}[1]
    \STATE Choose a fresh uniformly random integer $p \in [1, 2^{n}]$.

    \STATE If $p$ is prime, output $p$.  Otherwise start over.
  \end{algorithmic}
\end{algorithm}

Clearly, if $\algo{GenPrime}$ terminates, it outputs an element of
$\Pi_{n}$.  Moreover, it outputs a \emph{uniformly random} element of
$\Pi_{n}$, because every $\bar{p} \in \Pi_{n}$ is equally likely to be
chosen in the first step.  So the only remaining question is the
running time: can we test primality efficiently?  And can we guarantee
that the algorithm terminates quickly, i.e., does not start over too
many times?

The first question has many answers.  The definitive answer was
provided in 2002 by Agrawal, Kayal, and Saxena, who gave a
\emph{deterministic} primality test that runs in time polynomial in
the bit length of the number (i.e., $\poly(n)$ for a number in
$[1,2^{n}]$).  The running time, while polynomial, is quite large, so
the AKS test is not used in practice.  Instead, \emph{randomized}
tests such as Miller-Rabin are used to quickly test primality;
however, such tests have a \emph{very small} probability of returning
the wrong answer (``prime'' when the number is composite, or vice
versa, or both).  This introduces a tiny bias into the output
distribution of $\algo{GenPrime}$, which we can fold into a negligible
error term that we won't need to worry about.\footnote{Another
  alternative is to use \emph{deterministic} primality tests which
  only have proofs of correctness under unproved (but widely-believed)
  number-theoretic assumptions.  These are less common in practice
  because they aren't so fast.}  (Later on we will rigorously define
what ``negligible error term'' means.  We may study primality tests in
more detail.)

The second question, about the running time of $\algo{GenPrime}$, is
also very interesting.  The probability that a particular iteration
terminates is exactly $\abs{\Pi_{n}} / 2^{n}$ (assuming a perfect
primality test).  How does $\abs{\Pi_{n}}$ compare to $2^{n}$, i.e.,
how ``dense'' are the $n$-bit primes?  It is easy to show that there
are an infinite number of primes, but this is not strong enough for
our purposes --- primes could be so vanishingly rare that
$\algo{GenPrime}$ effectively runs forever.  Fortunately, it turns out
that the primes are quite dense.  For an integer $N > 1$, let $\pi(N)$
denote the number of primes not exceeding $N$.  We have the following:

\begin{lemma}[Chebyshev]
  \label{lem:chebyshev-primes}
  For every integer $N > 1$, $\pi(N) > \frac{N}{2 \log_{2} N}$.
\end{lemma}
(In fact, this is a weaker form of the ``prime number theorem,'' which
says that $\pi(N)$ asymptotically approaches $\frac{N}{\ln N}$.  We
will not give a proof of Lemma~\ref{lem:chebyshev-primes} now, but
there is an elementary one in the Pass-shelat notes.)  Using
Lemma~\ref{lem:chebyshev-primes}, we have \[
\frac{\abs{\Pi_{n}}}{2^{n}} = \frac{\pi(2^{n})}{2^{n}} >
\frac{2^{n}}{2^{n} \cdot 2 \log_{2}(2^{n})} = \frac{1}{2n}. \]
Finally, since all the iterations of $\algo{GenPrime}$ are
independent, the probability that it fails to terminate after (say)
$2n^{2}$ iterations is \[ (1-\tfrac{1}{2n})^{2n \cdot n} \leq
(1/2)^{n}, \] which is exponentially small.\footnote{To be completely
  formal, in order to say that $\algo{GenPrime}$ is polynomial-time,
  we should modify it so that if it has not terminated after $2n^{2}$
  iterations, it just outputs some fixed value (say, $1$).  This
  introduces only an exponentially small bias in its output
  distribution.}

\subsection{Weak OWFs and Hardness Amplification}
\label{sec:hardn-ampl}

The second way of dealing with the above difficulty of ``somewhere
hard'' functions is more general, but also most costly and technically
more complex.  Suppose we were not confident where the ``hard'' inputs
to a OWF were likely to be, but we did believe that they weren't too
rare.  We would then have a $\delta$-\emph{weak} one-way function,
where the ``hard to invert'' property from Definition~\ref{def:owf} is
relaxed to the following:
\begin{itemize}
\item For any non-uniform PPT algorithm $\Inv$, \[ \advan_{f}(\Inv) :=
  \Pr_{x \gets \bit^{n}} \bracks*{\Inv(f(x)) \in f^{-1}(f(x))} \leq
  1-\delta, \] for some $\delta = \delta(n) = 1/\poly(n)$.
\end{itemize}
The only change is that we have replaced the $\negl(n)$ success
probability by $1-\delta$.  This definition allows for $f$ to be
easily invertible for a large fraction of inputs (e.g., all the even
numbers), but not ``almost everywhere:'' a $\delta$-fraction of the
inputs remain hard.

Suppose that $f$ is a $\delta$-weak OWF.  Can we get a (strong) OWF
from it?  The answer is yes; this kind of result lies within the area
called \emph{hardness amplification} (for obvious reasons).  Assume
that we have a function $f$; then define its $m$-wise \emph{direct
  product} \[ f'(x_{1}, \ldots, x_{m}) = (f(x_{1}), \ldots,
f(x_{m})), \] where we parse the input so that $\len{x_{1}} = \cdots
= \len{x_{m}}$.  We have the following theorem:

\begin{theorem}
  \label{thm:weak-strong-owf}
  If $f$ is a $\delta$-weak one-way function, then $f'$ is a (strong)
  one-way function for any $\poly(n)$-bounded $m \geq 2n/\delta$.
\end{theorem}

\textbf{Proof idea:} The intuition here is that inverting $f'$
requires inverting $f$ on $m$ independently chosen values.  If each
component is inverted independently with probability at most
$1-\delta$, then the probability of inverting all of them should be at
most \[ (1-\delta)^{2n/\delta} \approx (1/e)^{2n} = \negl(n). \]
Rigorously proving that this intuition is correct, however, turns out
to be quite subtle.  The reason is that an adversary $\Inv'$ attacking
$f'$ does not have to treat each of the inversion sub-tasks $y_{1} =
f(x_{1}), \ldots, y_{m} = f(x_{m})$ independently.  For example, it
might look at all of the $y_{i}$s at once, and either invert
\emph{all} of them (with some non-negligible probability overall), or
none of them.  So the individual events (that $\Inv'$ succeeds in the
sub-task of inverting $y_{i} = f(x_{i})$) can be highly correlated.

To give a correct proof of the theorem, we need to perform a
\emph{reduction} from inverting $f$ with advantage $> 1-\delta$ to
inverting $f'$ with some non-negligible advantage.  Unpacking this a
bit, we want to prove the contrapositive: suppose there is some
(efficient) adversary $\Inv'$ that manages to break the (strong)
one-wayness of $f'$, i.e., it inverts with some non-negligible
advantage $\advan_{f'}(\Inv') = \alpha = \alpha(n)$.  Then we will
construct another (efficient) adversary $\Inv$ that manages to break
the $\delta$-weak one-wayness of $f$, i.e., it inverts $f$ with
advantage $\advan_{f}(\Inv) > 1-\delta$.  In doing so, we will use
$\Inv'$ as a ``black box,'' invoking it on inputs of our choice and
using its outputs in whatever way may be useful.  However, we cannot
assume that $\Inv'$ ``works'' in any particular way; we can only rely
on the hypothesis that it violates the strong one-wayness of $f'$.

The basic idea of the reduction is this: $\Inv$ is given $y = f(x)$
and wants to use $\Inv'$ to invert $y$.  An obvious first attempt is
to ``plug in'' the value $y$ as part of the output of $f'$, i.e., to
run $\Inv'$ on \[ y' = (y_{1} = y = f(x), y_{2} = f(x_{2}), \ldots,
y_{m} = f(x_{m})) = f'(x_{1} = x, x_{2}, \ldots, x_{m}) \] for some
random $x_{2}, \ldots, x_{m}$ chosen by $\Inv$.  When $\Inv'$ produces
a (candidate) preimage $x'=(x'_{1},\ldots,x'_{m})$, then $\Inv$ just
outputs $x'_{1}$ as its candidate preimage of $y=f(x)$.  Notice that
if $\Inv'$ happens to invert $y'$ successfully, then in particular
$x_{1} \in f^{-1}(y)$ as required.

Unfortunately, this simple idea does not quite work.  The problem is
that $\Inv'$ inverts only with non-negligible probability $\alpha$,
whereas we need $\Inv$ to invert with (typically larger) probability
$> 1-\delta$, to violate the weak one-wayness of $f$.  We might hope
to fix this problem by running $\Inv'$ a large number of times, using
the same $y_{1} = f(x)$ but \emph{fresh} random $x_{2}, \ldots, x_{m}$
each time, expecting $\Inv'$ to successfully invert at least once.
(Note that we can check whether $\Inv'$ succeeded in inverting $f'$ on
a particular tuple, because $f'$ is efficiently computable.)  This
strategy is better, but still flawed: the problem is that the runs are
\emph{not independent}, since we use the same $y_{1} = f(x)$ every
time.  Indeed, $\Inv'$ might behave in a ``sneaky'' way to make this
very strategy fail.  For example, $\Inv'$ might fix a certain
$\alpha$-fraction of ``good'' values of $x_{1}$ for which it always
inverts (ignoring $x_{2}, \ldots, x_{m}$), and fail on \emph{all}
other values of $x_{1}$.  Notice that such $\Inv'$ has overall
advantage $\alpha$, as required.  But no matter how many time we
repeatedly run $\Inv'$ with $y_{1}=y=f(x)$, we will eventually succeed
only if the original $x$ is ``good,'' which happens with probability
only $\alpha$.

A better reduction (which will turn out to work) just applies the
above ``plug-and-repeat'' strategy for \emph{each} position $i = 1,
\ldots, m$.  As it turns out, we can show that \emph{some} position
must have a very large ($> 1-\delta/2$) fraction of ``good''
$x_{i}$'s, for a suitable definition of good.  Conditioned on the $x$
from our given instance $y = f(x)$ being good, by repeatedly calling
$\Inv'$ a polynomial number of times (using fresh $x_{2}, \ldots,
x_{m}$ each time), the probability that $\Inv'$ inverts successfully
in at least one call will be extremely close to 1, so our overall
advantage in inverting $f$ will be $\approx 1-\delta/2 > 1-\delta$, as
desired.

We now give the formal proof.

\begin{proof}[Proof of Theorem~\ref{thm:weak-strong-owf}]
  Because $m = \poly(n)$ by assumption (and because $2n/\delta = 2n
  \cdot \poly(n) = \poly(n)$), if $f$ is efficiently computable, then
  so is $f'$.  Now suppose that some efficient adversary $\Inv'$
  violates the one-wayness of $f'$, with non-negligible advantage
  $\alpha = \alpha(n) = \advan_{\Inv'}(f')$.  We wish to construct an
  efficient adversary $\Inv$ that uses $\Inv'$ (efficiently) to
  violate the $\delta$-weak one-wayness of $f$.

  We first establish a crucial property of $\Inv'$ that will help us
  obtain the desired reduction.  Define the ``good'' set for position
  $i$ to be the set of all $x_{i}$ such that $\Inv'$ successfully
  inverts with ``good enough'' probability, over the random choice of
  all the remaining $x_{j}$.  Formally, \[ G_{i} = \set*{ x_{i}\;
    :\; \Pr_{x_{j} \text{ for } j \neq i} \bracks*{ \Inv'(f'(x'))
      \text{ succeeds} } \geq \frac{\alpha}{2m}}. \] Note that because
  $\Inv'$ succeeds with at least $\alpha$ probability overall, it is
  easy to see that $G_{i}$ is at least $(\alpha/2)$-dense for every
  $i$.  However, we will show a very different fact: that $G_{i}$ is
  at least $(1-\delta/2)$-dense for \emph{some} $i$.  (Note that
  typically, $(1-\delta/2)$ is much larger than $\alpha$.)

  \begin{claim}
    \label{claim:good-dense}
    For some $i$, we have $\frac{\abs{G_{i}}}{2^{n}} \geq
    (1-\delta/2)$.
  \end{claim}

  \begin{proof}
    Suppose not.  Then we have
    \begin{align*}
      \lefteqn{\Pr_{x'}[\Inv'(f'(x')) \text{ succeeds]}} \\
      &\leq \Pr[\Inv' \text{ succeeds} \wedge \text{every } x_{i} \in
      G_{i}] + \sum_{i=1}^{m} \Pr[\Inv' \text{ succeeds} \wedge
      x_{i} \not\in G_{i}] & \text{(union bound)} \\
      &\leq \Pr[\text{every } x_{i} \in G_{i}] + \sum_{i=1}^{m}
      \Pr[\Inv' \text{ succeeds}\; |\; x_{i} \not\in
      G_{i}] & \text{(probability)} \\
      &< (1-\tfrac{\delta}{2})^{2n/\delta} + m \cdot \tfrac{\alpha}{2m}
      & \text{(def \& size of $G_{i}$)} \\
      &\leq 2^{-n} + \alpha/2 < \alpha,
    \end{align*}
    contradicting our assumption on $\Inv'$.
  \end{proof}

  The remainder of the proof is a simple exercise, following the
  ``plug-and-repeat'' strategy described above, and using
  Claim~\ref{claim:good-dense}.
\end{proof}

\end{document}

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: t
%%% End: 
