\documentclass[11pt]{article}

\input{header}

\usepackage{algorithm,algorithmic}

% VARIABLES

\newcommand{\lecturenum}{4}
\newcommand{\lecturetopic}{Number Theory, OWF Variants}
\newcommand{\scribename}{Shiva Kintali}

% END OF VARIABLES

\lecheader

\pagestyle{plain}               % default: no special header

\begin{document}

\thispagestyle{fancy}           % first page should have special header

% LECTURE MATERIAL STARTS HERE

Today we will see some concrete one-way function candidates that arise
from number theory, and abstract out some of their other special
properties that will be useful when we proceed to investigate
pseudorandomness.

\section{Collections of OWFs}
\label{sec:coll-owfs-vari}

Our generic definition of a one-way function is concise, and very
useful for complexity-theoretic crypto.  However, it tends not to be
as appropriate for the kinds of hard functions that we use in
``real-life'' crypto; below we give a more flexible definition.  (In
your homework, you will show that the generic OWF definition is
equivalent to this one.)

\begin{definition}
  \label{def:collection-owfs}
  A \emph{collection of one-way functions} is a family $F = \set{
    f_{s} \colon D_{s} \to R_{s} }_{s \in S}$ satisfying the following
  conditions:
  \begin{enumerate}
  \item \emph{Easy to sample a function:} there is a PPT algorithm
    $\algo{S}$ such that $\algo{S}()$ outputs some $s \in S$
    (according to some arbitrary distribution).
  \item \emph{Easy to sample from domain:} there is a PPT algorithm
    $\algo{D}$ such that $\algo{D}(s)$ outputs some $x \in D_{s}$
    (according to some arbitrary distribution).
  \item \emph{Easy to evaluate function:} there is a deterministic
    polynomial-time algorithm $\algo{F}$ for which
    $\algo{F}(s,x)=f_{s}(x)$ for all $s \in S$, $x \in D_{s}$.
  \item \emph{Hard to invert:} for any non-uniform PPT algorithm
    $\Inv$, \[ \Pr_{s \gets \algo{S}(1^n), x \gets \algo{D}(s)}
    \bracks*{\Inv(s,f_{s}(x)) \in f_{s}^{-1}(f_{s}(x))} = \negl(n). \]
  \end{enumerate}
\end{definition}

For example, the subset-sum function $f_{\text{ss}}$ is more naturally
defined as a collection, as follows.  Let $S_{n} = (\Z_{N})^{n}$ where
$N=2^{n}$, and let the full index set $S = \cup_{n=1}^{\infty} S_{n}$.
Define the domain $D_{\vec{a}} = \bit^{n}$ and the range $R_{\vec{a}}
= \Z_{N}$, for all $\vec{a} = (a_{1}, \ldots, a_{n}) \in S_{n}$. The
corresponding function is defined as \[ f_{\vec{a}}(x) =
\sum_{i=1}^{n} a_{i} \cdot x_{i} \bmod N. \] The algorithms $\algo{S}$
(function sampler), $\algo{D}$ (domain sampler), and $\algo{F}$
(function evaluator) are all straightforward.

In the remainder of the lecture, we will see other examples of OWF
collections (some with other special properties) that arise from
number theory.

\section{Number Theory Background}
\label{sec:numb-theory-backgr}

\begin{definition}
   \label{def:gcd}
   For positive integers $a,b \in \N$, their \emph{greatest common
     divisor} $d = \gcd(a,b)$ is the largest integer $d$ such that $d
   \mid a$ and $d \mid b$.
\end{definition}

As a consequence of \cref{alg:exteuclid} below, there always
exist integers $x,y \in \Z$ such that $ax + by = \gcd(a,b)$.  We say
that $a$ and $b$ are \emph{co-prime} (or \emph{relatively prime}) if
$\gcd(a,b) = 1$, i.e., $ax = 1 \bmod b$.  From this, $x$ is the
multiplicative inverse of $a$ modulo $b$, and likewise $y$ is the
multiplicative inverse of $b$ modulo $a$.  The following deterministic
algorithm shows that $\gcd(a,b)$ (and additionally, the integers $x$
and $y$) can be computed efficiently.

\newcommand{\exteuclid}{\algo{ExtendedEuclid}}
\renewcommand{\algorithmicrequire}{\textbf{Input:}}
\renewcommand{\algorithmicensure}{\textbf{Output:}}

\begin{algorithm}
  \caption{Algorithm $\exteuclid(a,b)$ for computing the greatest
    common divisor of $a$ and $b$.}
  \label{alg:exteuclid}

  \begin{algorithmic}[1]
    \REQUIRE Positive integers $a \geq b > 0$.

    \ENSURE $(x,y) \in \Z^{2}$ such that $ax + by = \gcd(a,b)$.

    \IF {$b \mid a$}
    \RETURN $(0,1)$
    \ELSE
    \STATE Let $a = b \cdot q + r$ for $r \in \set{1, \ldots, b-1}$
    \STATE $(x',y') \gets \exteuclid(b,r)$
    \RETURN $(y', x' - q \cdot y')$
    \ENDIF
  \end{algorithmic}
\end{algorithm}

\begin{theorem}
  \label{thm:exteuclid}
  $\exteuclid$ is correct and runs in polynomial time in the
  \emph{lengths} of $a$ and $b$, i.e., in $poly(\log a + \log b)$
  time.
\end{theorem}

\begin{proof}
  For correctness, we argue by induction on the second argument $b$.
  Clearly the algorithm is correct when $b=1$.  If $b \mid a$, then
  $\gcd(a,b) = b$, hence $\exteuclid$ correctly returns $(0,1)$.  If
  $b \nmid a$ then by the inductive hypothesis (using $b > r$), the
  recursive call correctly returns $(x',y')$ such that $bx' + ry' =
  \gcd(b,r)$.  It can be checked that $\gcd(a,b) = \gcd(b,r)$, because
  any common divisor of $a$ and $b$ is also a divisor of $r$.
  Finally, observe that \[ \gcd(b,r) = bx' + ry' = bx' + (a - b \cdot
  q)y' = ay' + (x' - q \cdot y')b. \] Hence $\exteuclid$ correctly
  returns $(y', x' - q \cdot y')$.
  
  For the running time, observe that all the basic operations (not
  including the recursive call) can be implemented in polynomial
  time.  The following claim establishes the overall efficiency.
  
  \begin{claim}
    For $2^{n} > a \geq b > 0$, $\exteuclid$ makes at most $2n$
    recursive calls.
  \end{claim}

  We use induction.  The claim is true when $a < 2^{1}$.  Suppose the
  claim is true for all $a < 2^n$, and suppose $a < 2^{n+1}$.  Two
  cases arise:
  \begin{itemize}
  \item If $b < 2^n$, the first recursive call is on $(b, r)$.
    Since $b < 2^n$, by the inductive hypothesis we make at most
    $2n$ more recursive calls.  Hence the total number of recursive
    calls is at most $1 + 2n < 2(n+1)$.
  \item If $b \geq 2^n$, i.e., $2^{n+1} > a \geq b \geq 2^n$, we have
    $a = b \cdot 1 + r$ for $r = a-b < 2^n < b$.  The first recursive
    call is on $(b \geq 2^n, r < 2^n)$.  In turn, its recursive call
    uses $r < 2^n$ as its first parameter.  By the inductive
    hypothesis, the number of recursive calls following the second one
    is at most $2n$.  Hence the total number of recursive calls is at
    most $2 + 2n \leq 2(n+1)$.
  \end{itemize}
\end{proof}

\noindent We frequently work with the ring $(\ZN, +, \cdot)$ of
integers modulo a positive integer $N$.

\begin{lemma}[Chinese remainder theorem, special case]
  Let $N = p \cdot q$ for distinct primes $p,q$.  The ring $\Z_N$ is
  isomorphic to the product ring $\Zp \times \Zq$, via the isomorphism
  $h(x) = (x \bmod p, x \bmod q)$.
\end{lemma}

\noindent A few remarks about the above lemma:
\begin{itemize}
\item In the product ring $\Zp \times \Zq$, addition and
  multiplication are coordinate-wise.
\item Clearly the isomorphism $h$ is efficiently computable.  Less
  obvious is that it is also efficiently \emph{invertible}.  Suppose
  we know some elements $c_{p}, c_{q} \in \ZN$ such that
  $h(c_{p}) = (1,0)$ and $h(c_{q}) = (0,1)$; such a pair is sometimes
  called a \emph{CRT basis}.  Then given $(x,y) \in \Zp \times \Zq$,
  it is easy to see that
  $h^{-1}(x,y) = \bar{x} \cdot c_{p} + \bar{y} \cdot c_{q}$, where
  $\bar{x}, \bar{y} \in \ZN$ are any elements such that
  $\bar{x} = x \pmod{p}$ and $\bar{y} = y \pmod{q}$. (For example, we
  could take the ``smallest'' elements in~$\ZN$ that have the required
  residues modulo~$p$ and~$q$, respectively.)
\end{itemize}

\begin{question}
  For the CRT isomorphism on $\Z_{15}$, verify that the following
  example equations hold: $h(7 \cdot 9) = h(7) \cdot h(9)$ and
  $h(6 + 11) = h(6) + h(11)$.
\end{question}

\begin{answer}
  Note that $15$ is the product of two primes, $3$ and $5$, so
  \(\Z_{15} \cong \Z_3 \times \Z_5\).

  First, we consider \(h(7 \cdot 9)\). We have
  \(7 \cdot 9 = 3 \pmod{15}\) and
  \(h(3) = (3 \bmod 3, 3 \bmod 5) = (0, 3)\). Next,
  \(h(7) = (7 \bmod{3}, 7 \bmod{5}) = (1,2)\) and
  \(h(9) = (9 \bmod{3}, 9 \bmod 5) = (0, 4)\). Finally, we multiply
  the pairs elementwise, recalling that the first elements of each
  pair are from~\(\Z_3\) and the second elements are from~\(\Z_5\). We
  see that \((1,2) \cdot (0,4) = (0 \bmod 3, 8 \bmod 5) = (0,3)\), as
  expected.

  Now we consider \(h(6 + 11)\). We have \(6 + 11 = 2 \pmod{15}\) and
  \(h(2) = (2 \bmod 3, 2 \bmod 5) = (2,2)\). Next,
  \(h(6) = (6 \bmod{3}, 6 \bmod{5}) = (0,1)\) and
  \(h(11) = (11 \bmod{3}, 11 \bmod{5}) = (2,1)\). Finally, adding the
  pairs elementwise (and reducing modulo the appropriate moduli)
  yields \((0,1) + (2,1) = (2,2)\), as expected.
\end{answer}

\begin{question}
  Show that \(h^{-1}(x,y) = \bar{x} \cdot c_p + \bar{y} \cdot c_q\),
  as claimed above.
\end{question}

\begin{answer}
  We use the CRT isomorphism several times:
  \begin{align*}
    (x,y) &= (x, 0) + (0, y) \\
          &= h(\bar{x}) \cdot (1,0) + h(\bar{y}) \cdot (0,1) \\
          &= h(\bar{x}) \cdot h(c_p) + h(\bar{y}) \cdot h(c_q) \\
          &= h(\bar{x} \cdot c_p + \bar{y} \cdot c_q),
  \end{align*}
  and the result follows by applying the bijection~$h^{-1}$ to both
  sides.
\end{answer}

\begin{question}
  Show how to compute $c_{p}, c_{q}$ efficiently (hint: use
  $\exteuclid$ on $p,q$).
\end{question}

\begin{answer}
  Because $p,q$ are distinct primes, their gcd is 1, so
  $\exteuclid(p,q)$ returns integers $a,b \in \Z$ such that
  $a \cdot p + b \cdot q = 1$. We claim that
  $c_{p} = b \cdot q \bmod{p}$ and $c_{q} = a \cdot p \bmod{q}$. To
  see this, observe that $b \cdot q = 1 - a \cdot p = 1 \pmod{p}$
  (because~$a$ is an integer) and $b \cdot q = 0 \pmod{q}$
  (because~$b$ is an integer), so $h(b \cdot q) = (1,0)$, as needed. A
  similar calculation shows that $h(a \cdot p) = (0,1)$.
\end{answer}

\begin{definition}
  The multiplicative group $\ZN^{*} := \set{ x \in \Z_N : x \text{
      is invertible mod $N$, i.e., } \gcd(x,N) = 1 }$.
\end{definition}

\noindent Here are some useful facts about the multiplicative group
$\ZN^{*}$:
\begin{itemize}
\item For a prime $p$, $\Zp^{*} = \set{1, \ldots, p-1}$.
\item When $N=pq$ for distinct primes $p,q$, we have $\Z^{*}_N \cong
  \Zp^{*} \times \Zq^{*}$.
\end{itemize}

\begin{definition}
  For $N \in \Z^+$, Euler's \emph{totient function} $\varphi(N)$ is
  defined to be $\abs{\ZN^{*}}$, i.e., the number of positive integers
  $a \leq N$ that are relatively prime to $N$.
\end{definition}

\noindent Here are some useful facts about the totient function:
\begin{itemize}
\item For a prime $p$, we have $\varphi(p) = p - 1$.
\item For a prime $p$ and positive integer $a$, we have $\varphi(p^a)
  = (p-1) p^{a-1} = p^a - p^{a-1}$.
\item If $\gcd(a,b) = 1$, then $\varphi(a \cdot b) = \varphi(a) \cdot
  \varphi(b)$.
\end{itemize}

\begin{definition}
  The subgroup of \emph{quadratic residues} is defined as \[ \QRN^{*}
  = \set{y \in \ZN^{*} : \exists\; x \in \ZN^{*} \text{ s.t. } y = x^2
    \bmod N} \subseteq \ZN^{*}. \]
\end{definition}

\noindent Here are some useful facts about $\QRN^{*}$:
\begin{itemize}
\item For an odd prime $p$, $\abs{\QRp^{*}} = \frac{p-1}{2}$, because
  $x \mapsto x^{2}$ is $2$-to-$1$ over $\Zp^{*}$.  (Exercise: prove
  this.)

\item When $N = pq$ for distinct odd primes $p,q$, we have $\QRN^{*}
  \cong \QRp^{*} \times \QR_{q}^{*}$, hence $\abs{\QRN^{*}} =
  \frac{p-1}{2} \cdot \frac{q-1}{2}$.

\item For an odd prime $p$, we have $-1 \in \QRp^{*}$ if and only if
  $p = 1 \bmod 4$.
\end{itemize}

\begin{question}
    Verify that the CRT isomorphism works in $\Z_{15}$ by checking that $h(7 \cdot 9) = h(7) \cdot h(9)$.
\end{question}
\begin{answer}
    Note that $15$ is the product of two primes: $3$ and $5$, so \(\Z_{15} \cong \Z_3
    \times \Z_5\).  First, we consider \(h(7 \cdot 9)\). \(7 \cdot 9 \equiv 3 \pmod{15}\)
    and \(h(3) = (3 \bmod 3, 3 \bmod 5) = (0, 3)\). Next, \(h(7) = (7 \bmod{3}, 7
    \bmod{5}) = (1,2)\) and \(h(9) = (9 \bmod{3}, 9 \bmod 5) = (0, 4)\). Finally we
    multiply the pairs elementwise, recalling that the first elements of each pair are
    from \(\Z_3\) and the second elements are from \(\Z_5\).  \((1,2) \cdot (0,4) =
    (0,3)\) as expected.
\end{answer}
\begin{question}
    Show that \(h^{-1}(x,y) = x \cdot c_p + y \cdot c_q\).
\end{question}
\begin{answer}
    Consider that
    \begin{align*}
        (x,y) &= (x, 0) + (0, y) \\
              &= h(x) \cdot (1,0) + h(y) \cdot (0,1) \\
              &= h(x) \cdot h(c_p) + h(y) \cdot h(c_q) \\
              &= h(x \cdot c_p + y \cdot c_q) \\
        h^{-1}(x,y) &= x \cdot c_p + y \cdot c_q
    \end{align*}
\end{answer}

\section{Factoring-Related Functions}
\label{sec:fact-related}

We can abstract out a modulus generation algorithm $\algo{S}$, which
given the security parameter $1^{n}$ outputs the product $N$ of two
primes $p,q$.  For example, $\algo{S}$ might choose $p$ and $q$ to be
uniformly random and independent $n$-bit primes.

\emph{Rabin's function} $f_{N} \colon \ZN^{*} \to \QRN^{*}$ is defined
as follows: \[ f_{N}(x) = x^{2} \bmod N. \] Precisely defining the
collection according to \cref{def:collection-owfs} is a
simple exercise.  Note that $f_{N}$ is $4$-to-$1$, because each $y \in
\QRN^{*}$ has two square roots modulo $p$, and two modulo $q$.

\begin{theorem}
  \label{thm:rabin-owf-factor}
  If factoring is hard (with respect to $\algo{S}$), then the Rabin
  collection (with function generator $\algo{S}$) is one-way.
\end{theorem}

To prove the theorem, we will rely on the following essential number-theoretic fact.

\begin{claim}
  \label{claim:roots-factor}
  Let $N = pq$ be the product of distinct odd primes.
  Given any $x_{1}, x_{2} \in \ZN^{*}$ such that $x_{1}^{2} = x_{2}^{2} \bmod N$ but $x_{1} \neq \pm x_{2} \bmod N$, the factors of $N$ can be computed efficiently.
\end{claim}

\begin{proof}
  We have $x_{1}^{2} = x_{2}^{2} \bmod p$ and $x_{1}^{2} = x_{2}^{2}
  \bmod q$, which implies $x_{1} = \pm x_{2} \bmod p$ and $x_{1} =
  \pm x_{2} \bmod q$.  But we cannot have both $+$ or both $-$, by
  assumption.  Wlog, we have $x_{1} = + x_{2} \bmod p$ and $x_{1} =
  - x_{2} \bmod q$.  Then $p \mid (x_{1}-x_{2})$ but $q \nmid
  (x_{1}-x_{2})$, otherwise we'd have $q \mid (2x_{2}) \Rightarrow q
  \mid x_{2} \Rightarrow x_{2} \not\in \ZN^{*}$.  Then
  $\gcd(x_{1}-x_{2},N) = p$, which we can compute efficiently.
\end{proof}

\begin{proof}[Proof of \cref{thm:rabin-owf-factor}]
  First, as already discussed it is easy to generate a function,
  sample its domain, and evaluate the function.

  We prove one-wayness by contrapositive, via a reduction.
  Assuming we have an inverter for the Rabin function, the idea is to choose our own $x_{1} \in \ZN^{*}$ and invoke the inverter on $y = f_{N}(x_{1}) = x_{1}^{2} \bmod N$.
  The square root $x_{2}$ it returns will be $\neq \pm x_{1}$, with probability $1/2$.
  In such a case, we get the prime factorization of $N$ by \cref{claim:roots-factor}.
  We now proceed more formally.

  Assume a non-uniform PPT inverter $\Inv$ violating the one-wayness
  of the Rabin collection, i.e., \[ \Pr_{N \gets \algo{S}(1^{n}), x
    \gets \ZN^{*}}[\Inv(N, y = x^{2} \bmod N) \in \sqrt{y} \bmod N] =
  \delta(n) \] is non-negligible.

  Our factoring algorithm $\Adv(N)$ works as follows: first, generate
  a uniform $x_{1} \gets \ZN^{*}$.  Let $y = x_{1}^{2} \bmod N$ and
  let $x_{2} \gets \Inv(N, y)$.  If $x_{2}^{2} = y \bmod N$ but $x_{1}
  \neq \pm x_{2} \bmod N$, then compute the factorization of $N$ by
  \cref{claim:roots-factor}.

  We now analyze the reduction.  First, $N$ and $y$ are distributed as
  expected, so $\Inv$ outputs $x_{2}$ such that $x_{2}^{2} = y \bmod
  N$ with probability $\delta$.  Conditioned on the fixed value of
  $y$, there are four possible values for $x_{1}$, each equally likely
  by construction.  So we have $x_{2}^{2} = y \bmod N$ and $x_{2} \neq
  \pm x_{1} \bmod N$ with prob $\delta/2$, which is non-negligible by
  assumption.
\end{proof}

Suppose $p,q = 3 \bmod 4$.  Then $-1$ is not a square modulo $p$
(respectively, $q$).  So for any $x \in \Zp^{*}$ (resp., $\Zq^{*}$),
exactly one of $\pm x$ is a square modulo $p$ (resp., $q$).  From this
it can be seen that if we restrict the Rabin function to have domain
$\QRN^{*}$, i.e., $f_{N} \colon \QRN^{*} \to \QRN^{*}$, it becomes a
\emph{permutation} (bijection).

\medskip
\noindent {\bf{Question}}: Our proof that $f_{N}$ is one-way used
(quite essentially) the fact that $f_{N}$ is $4$-to-$1$.  Now that we
have changed its domain to make $f_{N}$ a permutation, is the proof
still valid?

\begin{definition}[One-Way Permutation]
  A collection $F = \set{f_{s} \colon D_{s} \to D_{s}}_{s \in S}$ is a
  collection of \emph{one-way permutations} if it is a collection of
  one-way functions $f_{s}$ under the \emph{uniform} distribution over
  $D_{s}$, and each $f_{s}$ is a \emph{permutation} of $D_{s}$ (i.e.,
  a bijection).
\end{definition}

\end{document}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:
